White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2015-09-08 papers/wdmerger_I/paper.tex
Point out that non-zero tolerances on gravity solve results in small deviations from conservation of energy.

Finally, then, we write the update in a zone for the total energy that conserves $(\rho E_{\text{tot}})$ as:
\begin{equation}
  \Delta (\rho E) = -\frac{1}{2}\sum_{f} \Delta \rho_{f} (\Phi_{f+1/2} - \Phi_{f-1/2}),\label{eq:grav_energy_conservation_update}
\end{equation}
where the sum is over the cell faces with indices $f$ and the indices $f+1/2$ and $f-1/2$ refer to 
the zone centers immediately to the left and right in the direction perpendicular to the face.
As long as we record the hydrodynamical fluxes through the zone faces after coming out of the hydrodynamics step, 
we are able to conserve the total energy completely (except for any energy loss or gain through 
physical domain boundaries). In order for the method to be second-order accurate in time, 
we need to use a time-centered $\Phi$ (which can be computed by averaging the time-level $n$ and $n+1$ potentials;
we already have the latter because \castro\ re-computes the potential at the new time after the hydrodynamics step,
and we can apply this energy at the end of the timestep). Note that of course the hydrodynamical
flux is already second-order accurate in time. In passing, we observe that the factor of $1/2$ that
appears in \autoref{eq:grav_energy_conservation_update} has nothing to do with the factor of
$1/2$ that appears in the statement of conservation of total energy, \autoref{eq:total_energy_gravity}.
The former comes simply from the fact that the energy change is evaluated using the mass motion through
a distance of half of the zone width. The latter is needed to ensure that these local changes in energy
are not double-counted when doing a global integral, since the gravitational potential is self-generated.
\autoref{eq:grav_energy_conservation_update} applies to any conservative potential $\Phi$, and we will
use this to our advantage for the rotation forces in \autoref{sec:rotation}.

>>>>

Finally, then, we write the update in a zone for the total energy that conserves $(\rho E_{\text{tot}})$ as:
\begin{equation}
  \Delta (\rho E) = -\frac{1}{2}\sum_{f} \Delta \rho_{f} (\Phi_{f+1/2} - \Phi_{f-1/2}),\label{eq:grav_energy_conservation_update}
\end{equation}
where the sum is over the cell faces with indices $f$ and the indices $f+1/2$ and $f-1/2$ refer to 
the zone centers immediately to the left and right in the direction perpendicular to the face.
As long as we record the hydrodynamical fluxes through the zone faces after coming out of the hydrodynamics step, 
this algorithm is able to conserve the total energy completely (except for any energy loss or gain through 
physical domain boundaries). In order for the method to be second-order accurate in time, 
we need to use a time-centered $\Phi$ (which can be computed by averaging the time-level $n$ and $n+1$ potentials;
we already have the latter because \castro\ re-computes the potential at the new time after the hydrodynamics step,
and we can apply this energy at the end of the timestep). Note that of course the hydrodynamical
flux is already second-order accurate in time. We observe also that in practice we will not obtain 
conservation of energy to machine precision even in the absence of open domain boundaries. The 
method itself is conservative if it is time-centered and correctly evaluates the energy change 
on cell faces \citep{marcello:2012}. However, in practice there is a non-zero numerical tolerance associated 
with the Poisson gravity solver (in our case, the multigrid method) that results in a non-zero error 
in the calculation of the gravitational potential. This results in a very small deviation from perfect 
conservation. It is not usually larger than the other effects which result in energy non-conservation 
for our simulations, such as resetting the state of zones that acquire a negative internal energy, and 
in principle if desired it can be made smaller by using stricter tolerance levels on the gravity solve.

In passing, the factor of $1/2$ that appears in \autoref{eq:grav_energy_conservation_update} 
has nothing to do with the factor of $1/2$ that appears in the statement of conservation of total energy, 
\autoref{eq:total_energy_gravity}. The former comes simply from the fact that the energy change is 
evaluated using the mass motion through a distance of half of the zone width. The latter is needed 
to ensure that these local changes in energy are not double-counted when doing a global integral, 
since the gravitational potential is self-generated. \autoref{eq:grav_energy_conservation_update} 
applies to any conservative potential $\Phi$, and we will use this to our advantage for the 
rotation forces in \autoref{sec:rotation}.
